The role of seed-dispersing animals in shaping plant distributions

### 2023-07-31 Text/Article_ModelSeedDispersal.Rmd
Update before submission

---
title: "The role of animal cognition in shaping resource distribution"
# "Animal cognition as a cause and consequence of resource distribution heterogeneity"
# "Eco-evolutionary feedback loops between animal cognition, movement and resource distribution"
# The role of animal cognition in shaping resource distribution via zoochory
# date: "`r format(Sys.time(), '%B %d, %Y')`"

@@

output: 
 #      - '--lua-filter=scholarly-metadata.lua'
 #      - '--lua-filter=author-info-blocks.lua'
 bookdown::pdf_document2:
   number_sections: true
   toc: false
   citation_package: biblatex
   latex_engine: pdflatex
   fig_caption: true
   pandoc_args:
     - '--lua-filter=../scholarly-metadata/scholarly-metadata.lua'
     - '--lua-filter=../author-info-blocks/author-info-blocks.lua'
always_allow_html: true

@@

**Abstract |** In the wild, an evolutionary play takes place in an unchanging setting, or so current hypotheses on the evolution of spatiotemporal cognition of animals implicitly assume. If plant-eating vagile animals are likely to have their spatiotemporal cognition shaped by the distribution of plant resources, their movement may shape the way resources are distributed as well, by dispersing seeds. This could create eco-evolutionary feedback loops between animal spatiotemporal cognition and the distribution patterns of plant resources. Using an agent-based model simulating the foraging behaviour of a seed disperser endowed with spatiotemporal knowledge of resource distribution, I investigated whether resource spatiotemporal patterns could be influenced by the level of cognition involved in foraging. This level of cognition represented how well resource location and phenology were predicted by the agent. I showed that seed dispersers could shape the long-term resource distribution with the formation of materialised resource routes. This stemmed from the conjunction of two forces: competition for space between trees and a seed-dispersing forager moving from tree to tree based on spatiotemporal memory. In turn, this engineering affected the benefits of spatiotemporal memory, hence the selection gradient that applies to cognition. Altogether, the results emphasise that foraging cognition is a cause and a consequence of resource heterogeneity.

@@

\initial{\textcolor{black}{H}}ypotheses about the evolution of vagile plant-eating foragers' spatiotemporal cognition (i.e. the ability to process, integrate and utilise information on resource distribution and phenology) generally assume that foragers evolve in an unchanging environment [@rosati2017foraging]. However, while foragers' cognition is likely to be shaped by the distribution of resources [@milton1981distribution; @grove2013evolution; @robiratemporalmemory; @bracis2015memory; @boyer2010modelling], the characteristics of the resources themselves could be the result of forager-induced selection [@lomascolo2010signal; @rojas2021combination]. In the tropical rainforests, for example, the cognition of frugivorous primates has probably been shaped by the challenge of the ephemerality and dispersal of their food [@milton1981distribution; @zuberbuhler2010foraging; @trapanese2019review], supporting complex [@teichroeb2019navigation; @trapanese2019whatwhere; @janson2016capuchins; @janson1998experimental; @janmaat2013mistake; @robirarandomwalk] and flexible [@trapanese2022flexibility; @janmaat2006primates] movement heuristics. In turn, primates remember and forage for the most rewarding trees [@ban2016fattyfruit; @ban2014highreward; @florchinger2010fruit]. The latter are thus more likely to have their seeds dispersed, driving plant and landscape evolution [@chapman2013primates; @lambert1998evolutionary].

|   The way seeds are dispersed contributes to the emerging pattern of resource distribution [@vittoz2007seed]. Most of the world's plants [@tiffney1995angiosperm] and especially rainforest trees [@gottsberger1983dispersal; @bagchi2011spatial] are dispersed by animals (zoochory). Compared to other modes of dispersal, such as wind (anemochory) or gravity (gravichory), zoochory induces a more patchy distribution of adult plants [@seidler2006seed]. This is because animals may tend to revisit memorised sites regularly, if not exclusively [@john2016memoryseed; @addis2017causes]. Furthermore, the imperfection and limited size of spatial memory [@fagan2013spatial] is expected to lead to different movement patterns [@avgar2013empirically], and therefore potentially to different patterns of resource distribution [@cortes2013integrating]. By being affected by cognition and by affecting resource dispersal, animal movement could thus lead to eco-evolutionary feedback loops between animal cognition and resource dispersal [@borah2022studying]. In other words, foraging cognition could be both a cause and a consequence of resource heterogeneity.

|   To illustrate the possible retroactions between cognition and resource distribution, let's consider an extreme speculative scenario in which seed-dispersing frugivores are freely foraging (Figure \@ref(fig:predictions)). At start, assuming a parcimonious homogeneous distribution, long-term spatiotemporal memory is only marginally advantageous [@grove2013evolution]. As it is nonetheless, cognition should be selected for. Through zoochory, deliberate targeting of some resource patches over others may induce a change in resource distribution over time (e.g. increasing patchiness), eventually making cognition more and more advantageous. Gradually, by repeatedly targeting the same trees, the routes used by the foragers may begin to materialise through the seeds scattered along the way. However, in doing so, foragers who can only passively follow these routes and have survived may also begin to become very efficient: they should not even have to bear the cost of high cognitive ability, putting them at an advantage over foragers who bear the burden of cognition [@raichle2006brain; @burns2011costs]. While the pattern of resource distribution can be maintained because the routes are used equally by all foragers [but through two different mechanisms: long-term memory *vs* taxis, @benhamou1995comparative; @benhamou1992distinguishing], cognition could even start to be counter-selected.

|   While some facets of this scenario may be rooted in empirical evidence [e.g. with forest elephants, @campos2011megagardeners; @blake2004fruit; @blake2009forest], most of it remains speculative. The aim of this study is thus to put such a scenario to the test. Using an agent-based model simulating the movement of seed-dispersing frugivores with different cognitive abilities, I sought to investigate (a) whether and how the spatial arrangement of resource varies with the cognitive level and movement behaviour of the simulated agents, and (b) whether and how this in turn affects the selection pressure on the agents' cognition.

@@

<!-- Yet, despite such a bipartite framework, with animal and (plant) resource, current research discarded evolution of both side. Most papers actually discard dynamism (e.g. mine), some consider competition between individuals but static memory (e.g. louise), and some consider dynamic memory (e.g. associative memory royal society paper lindt I think etc.). Yet, an animal might act upon its own environment too. -->

@@

The main objective of the model was to simulate the foraging behaviour of a single agent in an environment dynamical both on the short term, through an interplay between resource growth and depletion, and on the long term, due to seed dispersal. For simplicity, I will refer to the agent as the forager, and to resource patches as fruit trees, but this model should find larger resonance. The model was implemented in the *Rcpp* language in *R* software [`r extractRversion()`, @Rsoftware]  based on the *Rcpp* package [@Rcpp1; @Rcpp2; @Rcpp3].

@@

Time was cyclical, with a period of `r cycleLength` arbitrary time units (tu). Each tree could produce resource (fruit containing seeds) for `r fruitingLength` tu only once every period. Thus, resource was seasonal at the tree-level, but not at the map-level. Periods of productivity of each tree were randomly distributed (i.e. uniform distribution of the start date). Productivity was characterised by a triangular-shaped pattern of the quantity of food along time: at start of the productive period, food quantity started to increase linearly up to `r maxFoodToYield` arbitrary food unit (fu), mimicking progressive fruit ripening, and then decreased linearly to 0 fu, mimicking fruit rotting. At a given time, the quantity of food available at a given tree thus depended on the productivity of the tree at that time, as well as the depletion history by the agent, creating a short-term dynamics of resource temporal distribution.

@@

Trees could reproduce and disperse via fruit eaten by the agent. Once the agent ate on ripe fruit, seeds could be dispersed once up to `r timeDelayForDispersal` tu after ingestion, with a probability of `r DispersalProbability/timeDelayForDispersal` tu$^{-1}$, independent of the quantity of fruit ingested, for simplicity. Thus, all fruit eaten did not mandatorily led to dispersal. Dispersal was eventually successful only if some space was left available for the new tree: no tree had to be at a distance lower than `r round(0.45 * sqrt(mapSize * mapSize / (numberTrees * pi)), digits = 2)` su. This number was obtained such as the cumulative (when non overlapping) exclusive "lifespace" by trees (a circular area of radius the aforementioned distance) reached at best 45% of the map area (but other percentages were tested, see [Route formation as a product of two forces][Q3. Route formation as a product of two forces? Tree competition for space and forager movement step rule]). If it happened that there was available space for the seedling to grow, then the seedling location was sampled in the available space along the linear path linking the previous and current locations. Otherwise, the seedling could not grow and died.

@@

|   If the seedling was successfully dispersed and lived, an old tree (already present in the map) was randomly selected and died to keep density constant. The start date of fruiting of the new-born was randomly sampled within a reduced Gaussian distribution (i.e. variance of one) centered on the start date of fruiting of the parent tree. The location and fruiting timing of the new-born tree were memorised if it replaced a tree that was part of the agent's memory. The new-born tree was considered as mature only the year after. Seed dispersal could thus induce a long-term dynamics of resource spatial distribution and of the distribution of starting dates in the tree population.

@@

The forager was endowed with a gradient of knowledge abilities: from no memory of the locations of trees (naive forager) and the timing of their period of productivity to perfect omniscience (i.e. all tree locations are known) and prescience (i.e. all periods of productivity are known). In case a tree was part of the forager's memory, the forager knew both the spatial (location) and temporal (fruiting timing) attributes. Thus, it could not know only one or the other. In case of partial knowledge, known trees were randomly selected at start.

@@

The forager was initially randomly located within the map. It is fully aware of trees present within sensory range fixed to `r round(perceptualRange, digit=2)` su. This actually corresponds to the average nearest-neighbour distance which equates $\frac{1}{2\sqrt{\rho}}$, with $\rho$ the density of trees (here `r numberTrees/mapSize/mapSize`) at start of the simulation where trees are homogeneously distributed. It is also aware of trees part of its spatiotemporal long-term memory. To decide where to head next, the forager could either rely on its private knowledge (sensory or memory) or not. If all trees within its private knowledge did not yield any fruit (i.e., all having less than 0.001 fu), the forager was moving randomly. Random movements were performed by randomly choosing a direction (uniform distribution of the heading angle), while the step length was sampled within an exponential distribution of mean $\lambda_{step\,length}$ = `r exponentialRate`. An exponential distribution of step lengths is indeed classically observed in nature [@benhamou2014scales]. Otherwise, the agent targeted the `r whatRule` tree which maximised the ratio between the predicted food quantity at the tree (which accounts for depletion history), and the distance to the tree. Nonetheless, the forager avoided to target trees recently visited within the previous `r noReturnTime` tu to avoid targeting recently depleted trees [@robiratemporalmemory].

|   Once the target had been chosen (or random point, if none), the forager moved straight to the first tree that should have been theoretically perceived *en route* to the initial destination point if any (provided they were not recently visited), or the destination otherwise. When this destination is reached, it then re-estimated the best tree to target and so on. After each moving bout time was updated based on the distance travelled, knowing that the forager moved at a speed of `r speed` su/tu. Then, the resource quantity available at each tree was updated.

@@

To assess whether different levels of cognition resulted in different organisations of the spatial distribution of trees at the end of the simulation I focused on three metrics describing the tree spatiotemporal patterns: the *patchiness*, the *alignment* and the *spatial autocorrelation of fruiting timing*.

@@

|   As the forager is set up to forage in a square map with reflective boundaries, seed dispersal may lead to the shrinkage of resource range towards the centre of the map (where recursions are inevitably the higher), an artefact not of interest there. This would contribute to inflate measures of patchiness. I therefore calculated a measure of shrinkage *s* as one minus the area covered by the 95% of the kernel distribution based on all tree locations divided by the map size. To account for variations in shrinkage in the different simulated scenarios [(see Appendix Material)][Variations in resource range shrinkage intensity], I corrected patchiness such as $P_c = P(1-s)$.

@@

Routes are (curvi-)linear features. To assess the tendency of trees to form linear motifs, I quantified their alignment. For a given tree, alignment *a* was defined as one minus the minimum of the sine of the angles of the triangle formed by the tree of interest and the two closest trees. This varies from perfect alignment (value of 1), to no alignment (i.e., the minimum angle would be at worse 60° hence an alignment of ca. 0.134). Therefore, I used a corrected alignment $a_c$ varying from 0 (no alignment) to 1 (perfect alignment) such as

@@

Spatial autocorrelation of fruiting timing (i.e. the tendency of fruiting date of trees to more closely match that of neighbours than of distant trees) was calculated as the Geary Index, \textit{C} [@geary1954contiguity] applied to the start of fruiting dates. Since the date of start of fruiting corresponds to a circular variable, I followed [@schmal2017moran] to adapt existing functions of the *ape* package [@ape] to circular variables. Geary's \textit{C} was therefore defined as

@@

$C = \frac{(N-1)}{2} \frac{\sum_{i}\sum_{j} w_{ij}d(t_i,t_j)^2}{\sum_{i,j}w_{ij}\sum_i d(x_i,\bar{t})^2}$

@@

where $N$ is the number of trees, $t$ is the fruiting date (in radian), $d(t_1,t_2) = atan2(sin(t_1 - t_2), cos(t_1 - t_2))$, $\bar{t} = atan2(1/N\sum_isin(t_i), 1/N\sum_icos(t_i))$. I used the inverse of the Euclidean distance between tree locations as weights. Values of $C$ close to 1 indicate absence of spatial autocorrelation. Values close to 0 indicate negative spatial autocorrelation. Values larger than 1 indicate positive spatial autocorrelation.

@@

For each simulation, I verified whether a routine behaviour emerged and paralleled the variations in tree distribution. I characterised routine holding on the notion of conditional entropy following [@riotte2016randomness] using the series of the id of trees targeted during five supplementary seasons in which dispersal could not occur. I computed the routine possibly up to only a first order conditional entropy, as computing routine indices is computationally demanding and a first order is expecting to be the most informative to describe a perfect traplining behaviour [@riotte2016randomness]. The routine index varies from 0 (no routine) to 1 (perfect traplining).

@@

To investigate whether different levels of cognition eventually led to different distribution patterns, I considered five forager types differing by the proportion of trees for which the location and phenology was known (from 0 to 100%, at a step of `r 100/length(spatialKnowledge)`, henceforth *spatiotemporal knowledge rate*). Each forager was first let foraging individually for `r cycleLimitNumber` seasonal periods. I repeated the simulations for a given condition (i.e. knowledge rate) `r numberRepetitions` times.

@@

To investigate whether the engineered distribution of trees could in turn affect the foraging benefices purported by cognition, I considered the final tree distributions of simulations computed to answer [question 1][Q1. Does resource spatiotemporal distribution vary with cognition level?]). For each of these tree distributions, in three different simulation runs, I let forage a forager endowed with no, intermediate (i.e. half of trees known), or full (all trees known; omniscience) spatiotemporal knowledge. In these simulations, no seed-dispersal could occur. I measured the foraging efficiency after the forager foraged for `r cycleLimitNumber/5` full seasons and compared them across environments and forager abilities. Foraging efficiency was computed as the cumulated quantity of eaten food, over the cumulated distance travelled. I verified that at the end of the simulation that foraging efficiency stabilised, which meant that I reached a foraging equilibrium. I considered that foraging efficiency stabilised when the foraging efficiency sampled after each moving bout during the fourth and fifth quintile of simulation times did not statistically differed of more than `r percentToAssumeDifference`% (reference is the minimum efficiency in the fourth or the fifth quintile of simulation times). Distributions of foraging efficiency measures were compared with Student tests ("t.test" function). I repeated the simulations for a given conditions (i.e. knowledge rate) `r numberRepetitions` times.

@@

During the model construction, I noticed that routes eventually appeared with the implementation of two mechanisms: space competition between trees and sequential targeting of all trees seen *en route* to the target. To better quantify the effect of those two forces, I considered two other sets of `r numberRepetitions` simulations each.

|   To assess the effect of competition for space between trees, I compared the final tree spatiotemporal patterns when a perfectly knowledgeable forager (all tree locations and phenology known) was let foraging in the same setting as in question 1), with the only difference that the exclusive lifespace between trees could vary be of 5 (small lifespace), 45 (intermediate lifespace) or 85% (large lifespace) of the map surface, so that the radius of this exclusive circular lifespace was always inferior to the perception range (`r round(0.85 * sqrt(mapSize * mapSize / (numberTrees * pi)), digits = 2)` su vs `r round(perceptualRange, digit=2)` su).

@@

```{r figEngineering, fig.width=19.25, fig.height=19.25, include=TRUE, warning = FALSE, fig.cap="Seed-dispersing foragers can shape resource distribution | The plots show the results of 200 simulations when a seed-dispersing forager with varying level of cognition, moving choice rules or space competition intensities between trees (left to right) was let foraging for a long time. The resulting distribution pattern was assessed by its patchiness (the tendency of trees to be aggregated into patches; larger values indicate highly aggregated trees), its alignement (the tendency of trees to form linear geometries; larger values indicate the presence of linear geometries) and spatial autocorelation in fruiting date (the tendency of trees to have fruiting dates closer/farther in time to their spatial neighbours; values close to 0 indicate negative spatial autocorrelation, values close to 1 indicate no spatial autocorrelation, values above 1 indicate positive spatial autocorrelation). The emerging movement behaviour of the forager was described in terms of routine behaviour (the tendency of the forager to target visit trees in a predictible way; values close to 0 indicate no predictibly movement, values close to 1 indicate highly predictible movements)."}

@@

```{r figEfficiency, fig.width=8, fig.height=4, include=TRUE, warning = FALSE, fig.cap=" Relative foraging benefits of cognition are sensitive to changes in the emerging resource distribution | The 'initial condition' of resource distribution was taken from the simulations in which a null, intermediate or omniscient forager was first let foraging. Then, those three forager types were let foraging (separately) without dispersal and foraging efficiency was measured (cumulated food eaten over cumulated travelled distance)."}

@@

>>>>

---
title: "The role of seed-dispersing animals in shaping plant distributions"
# "Animal cognition as a cause and consequence of resource distribution heterogeneity"
# "Eco-evolutionary feedback loops between animal cognition, movement and resource distribution"
# The role of animal cognition in shaping resource distribution via zoochory
# date: "`r format(Sys.time(), '%B %d, %Y')`"

@@

output: 
 #      - '--lua-filter=scholarly-metadata.lua'
 #      - '--lua-filter=author-info-blocks.lua'
 bookdown::pdf_document2:
   number_sections: true
   toc: false
   citation_package: biblatex
   latex_engine: pdflatex
   fig_caption: true
   pandoc_args:
     - '--lua-filter=scholarly-metadata/scholarly-metadata.lua'
     - '--lua-filter=author-info-blocks/author-info-blocks.lua'
always_allow_html: true

@@

**Abstract |** In the wild, an evolutionary play takes place in an unchanging setting, or so current hypotheses on the evolution of spatiotemporal cognition of animals implicitly assume. However, if plant-eating vagile animals are likely to have their spatiotemporal cognition shaped by the distribution of plant resources, their movement may shape the way resources are distributed as well, for example by dispersing seeds. Using an agent-based model simulating the foraging behaviour of a seed disperser endowed with spatiotemporal knowledge of resource distribution, I investigated whether resource spatiotemporal patterns could be influenced by the level of cognition involved in foraging. This level of cognition represented how well resource location and phenology were predicted by the agent. I showed that seed dispersers could shape the long-term distribution of resources by materialising the routes repeatedly used by the forager with the newly recruited trees. This stemmed from the conjunction of two forces: competition for space between trees and a seed-dispersing forager moving from tree to tree based on spatiotemporal memory. In turn, resource landscape modifications affected the benefits of spatiotemporal memory, hence the selection gradient that applies to cognition. This could create eco-evolutionary feedback loops between animal spatiotemporal cognition and the distribution patterns of plant resources. Altogether, the results emphasise that foraging cognition is a cause and a consequence of resource heterogeneity.

@@

\initial{\textcolor{black}{H}}ypotheses about the evolution of vagile plant-eating foragers' spatiotemporal cognition (i.e. the ability to process, integrate and utilise information on resource distribution and phenology) generally assume that foragers evolve in an unchanging environment [@rosati2017foraging]. However, while foragers' cognition is likely to be shaped by the distribution of resources [@milton1981distribution; @grove2013evolution; @robiratemporalmemory; @bracis2015memory; @boyer2010modelling], the characteristics of the resources themselves could be the result of forager-induced selection [@lomascolo2010signal; @rojas2021combination]. In the tropical rainforests, for example, the cognition of frugivorous primates has probably been shaped by the challenge of the ephemerality and dispersion of their food [@milton1981distribution; @zuberbuhler2010foraging; @trapanese2019review], supporting complex [@teichroeb2019navigation; @trapanese2019whatwhere; @janson2016capuchins; @janson1998experimental; @janmaat2013mistake; @robirarandomwalk] and flexible [@trapanese2022flexibility; @janmaat2006primates] movement heuristics. In turn, primates remember and forage for the most rewarding trees [@ban2016fattyfruit; @ban2014highreward; @florchinger2010fruit]. The latter are thus more likely to have their seeds dispersed, driving plant and landscape evolution [@chapman2013primates; @lambert1998evolutionary].

|   The way seeds are dispersed contributes to the emerging pattern of resource distribution [@vittoz2007seed]. Most of the world's plants [@tiffney1995angiosperm] and especially rainforest trees [@gottsberger1983dispersal; @bagchi2011spatial] are dispersed by animals (zoochory). Compared to other modes of dispersal, such as wind (anemochory) or gravity (gravichory), zoochory induces a less patchy distribution of adult plants [@seidler2006seed; but see @hubbell1979tree]. However, animal-dispersed fruit trees are still not homogeneously distributed [@hubbell1979tree]. This is because although travelling long distances, animals may tend to revisit memorised sites regularly, if not exclusively [@john2016memoryseed; @addis2017causes]. Furthermore, the imperfection and limited size of spatial memory [@fagan2013spatial] is expected to lead to different movement patterns [@avgar2013empirically], and therefore potentially to different patterns of resource distribution [@cortes2013integrating]. By being affected by cognition and by affecting resource dispersal, animal movement could thus lead to eco-evolutionary feedback loops between animal cognition and resource dispersal [@borah2022studying]. In other words, foraging cognition could be both a cause and a consequence of resource heterogeneity.

|   To illustrate the possible retroactions between cognition and resource distribution, let's consider an extreme speculative scenario in which seed-dispersing frugivores are freely foraging (Figure \@ref(fig:predictions)). At start, assuming a random homogeneous distribution, long-term spatiotemporal memory can be advantageous and selected for [@grove2013evolution; @bracis2015memory]. Through zoochory, deliberate targeting of some resource patches over others may induce a change in resource distribution over time (e.g. increasing patchiness), eventually making cognition more and more advantageous [@grove2013evolution; @bracis2015memory; @boyer2010modelling]. Gradually, by repeatedly targeting the same trees, the routes used by the foragers may begin to materialise through the seeds scattered along the way. However, in doing so, foragers who can only passively follow these routes and have survived may also begin to become very efficient: they should not even have to bear the cost of high cognitive ability, putting them at an advantage over foragers who bear the burden of cognition [@raichle2006brain; @burns2011costs]. While the pattern of resource distribution can be maintained because the routes are used equally by all foragers [but through two different mechanisms: long-term memory *vs* taxis, @benhamou1995comparative; @benhamou1992distinguishing], cognition could even start to be counter-selected.

|   While some facets of this scenario may be rooted in empirical evidence [e.g. with forest elephants, @campos2011megagardeners; @blake2004fruit; @blake2009forest], most of it remains speculative. The aim of this study is to put such a scenario to the test. Using an agent-based model simulating the movement of seed-dispersing frugivores with different cognitive abilities, I sought to investigate (a) whether and how the spatial arrangement of resource varies with the cognitive level and movement behaviour of the simulated agents, and (b) whether and how this may in turn affect the selection pressure on the agents' cognition.

@@

<!-- Yet, despite such a bipartite framework, with animal and (plant) resource, current research discarded evolution of both side. Most papers actually discard dynamism (e.g. mine), some consider competition between individuals but static memory (e.g. louise), and some consider dynamic memory (e.g. associative memory ryal society paper lindt I think etc.). Yet, an animal might act upon its own environment too. -->

@@

The main objective of the model was to simulate the foraging behaviour of a single agent in an environment that is dynamical both on the short term, through an interplay between resource growth and depletion, and on the long term, due to seed dispersal. For simplicity, I will refer to the agent as the forager, and to resource patches as fruit trees, but this model should find larger resonance. The model was implemented in the *C++* language in *R* software [`r extractRversion()`, @Rsoftware]  based on the *Rcpp* package [@Rcpp1; @Rcpp2; @Rcpp3].

@@

Time was cyclical, with a period of `r cycleLength` arbitrary time units (tu). Each tree produced resource (fruit containing seeds) for `r fruitingLength` tu only once every period. Periods of productivity of each tree were randomly distributed (i.e. uniform distribution of the start date). Thus, resource was seasonal at the tree-level, but not at the map-level. Productivity was characterised by a triangular-shaped pattern of the quantity of food over time: at start of the productive period, food quantity started to increase linearly up to `r maxFoodToYield` arbitrary food unit (fu), mimicking progressive fruit ripening, and then decreased linearly to 0 fu, mimicking fruit rotting. The depletion of the food by the forager had no effect on this dynamic. Food could continue to ripen (though never to the same extent as if it had not been depleted) and rot. At a given time, the quantity of food available at a given tree thus depended on the productivity of the tree at that time, as well as the depletion history by the forager, creating a short-term dynamics of resource temporal distribution.

@@

Trees reproduced by ingestion of ripe fruit by the forager and subsequent dispersal. Once the forager ate ripe fruit, seeds could be dispersed only one time up to `r timeDelayForDispersal` tu after ingestion. Dispersal occurred at a probability of `r DispersalProbability/timeDelayForDispersal` tu$^{-1}$, independent of the quantity of fruit ingested. This probability was considered low, to mimic the natural slow pace of land use changes. All fruit eaten did not mandatorily led to dispersal. Dispersal was eventually successful only if some space was left available for the new tree: no tree had to be at a distance less than `r round(0.45 * sqrt(mapSize * mapSize / (numberTrees * pi)), digits = 2)` su. This number was obtained such that the cumulative (when non-overlapping) exclusive "lifespace" by trees (a circular area of radius the aforementioned distance) could reach 45% of the map area at most (but other percentages were tested, see [Route formation as a product of two forces][Q3. Route formation as a product of two forces? Tree competition for space and forager movement step rule]). If it happened that there was available space for the seedling to grow, then the seedling location was sampled in the available space along the linear path linking the previous and current locations. Otherwise, the seedling could not grow and died.

@@

|   If the seedling was successfully dispersed and recruited, an old tree (already present in the map) was randomly selected and died to keep density constant. The start date of fruiting of the newly recruited tree was randomly sampled within a reduced Gaussian distribution (i.e. variance of one) centered on the start date of fruiting of the parent tree. The location and fruiting time of the newly recruited tree were memorised if it replaced a tree that was part of the forager's memory. The newly recruited tree was considered as mature only the year after. Seed dispersal could thus induce a long-term dynamics of resource spatial distribution and of the distribution of starting dates in the tree population.

@@

The forager was endowed with a gradient of knowledge abilities: from no memory of the locations of trees (naive forager) and the timing of their period of productivity to perfect omniscience (i.e. all tree locations are known) and prescience (i.e. all periods of productivity are known). Otherwise, it only knew a certain proportion (fixed for the whole simulation) of tree locations and their associated phenology (hereafter *spatiotemporal knowledge rate*). These trees were selected randomly at the start of the simulation. When a tree was part of the forager's memory, the forager knew both the spatial (location) and temporal (fruiting timing) attributes. Thus, it could not know only one or the other.

@@

The forager was initially randomly located within the map. It is fully aware of trees present within sensory range fixed to `r round(perceptualRange, digit=2)` su. This actually corresponds to the average nearest-neighbour distance which equates $\frac{1}{2\sqrt{\rho}}$, with $\rho$ the density of trees (here `r numberTrees/mapSize/mapSize`) at start of the simulation where trees are homogeneously distributed. It is also aware of trees that are part of its spatiotemporal long-term memory. To decide where to head next, the forager could either rely on its private knowledge (sensory or memory) or not. If all trees within its private knowledge did not yield any fruit (i.e., all having less than 0.001 fu), the forager was moving randomly. Random movements were performed by randomly choosing a direction (uniform distribution of the heading angle), while the step length was sampled within an exponential distribution of mean $\lambda_{step\,length}$ = `r exponentialRate`. An exponential distribution of step lengths is indeed classically observed in nature [@benhamou2014scales]. Otherwise, the forager targeted the `r whatRule` tree which maximised the ratio between the predicted food quantity at the tree (which accounts for depletion history), and the distance to the tree. Nonetheless, the forager avoided to target trees recently visited within the previous `r noReturnTime` tu to avoid targeting recently depleted trees [@robiratemporalmemory].

|   Once the target had been chosen (or random point, if none), the forager moved straight to the first tree that should have been theoretically perceived *en route* to the initial destination point if any (provided they were not recently visited), or the destination otherwise. When this destination was reached, it then re-estimated the best tree to target and so on. After each moving bout time was updated based on the distance travelled, knowing that the forager moved at a speed of `r speed` su/tu. Then, the resource quantity available at each tree was updated.

@@

To assess whether different levels of cognition resulted in different organisations of the spatial distribution of trees at the end of the simulation, I focused on three metrics describing tree spatiotemporal patterns: the *patchiness*, the *alignment* and the *spatial autocorrelation of fruiting timing*.

@@

|   As the forager is set up to forage in a square map with reflective boundaries, seed dispersal may lead to the shrinkage of resource range towards the centre of the map (where recursions are inevitably higher), an artefact not of interest here. This would contribute to inflate measures of patchiness. I therefore calculated a measure of shrinkage *s* as one minus the area covered by the 95% of the Gaussian-based kernel distribution based on all tree locations divided by the map size. To account for variations in shrinkage in the different simulated scenarios [(see Appendix Material)][Variations in resource range shrinkage intensity], I corrected patchiness such as $P_c = P(1-s)$.

@@

Routes are (curvi-)linear features. To assess the tendency of trees to form linear motifs, I quantified their alignment. For a given tree, alignment *a* was defined as one minus the minimum of the sine of the angles of the triangle formed by the tree of interest and the two closest trees. This varies from perfect alignment (value of 1), to no alignment (i.e., the minimum angle would be at worst 60° hence an alignment of ca. `r round(1 - sin(pi/3), digits = 3)`). Therefore, I used a corrected alignment $a_c$ varying from 0 (no alignment) to 1 (perfect alignment) such as

@@

Spatial autocorrelation of fruiting timing (i.e. the tendency of fruiting date of trees to more closely match that of neighbours than of distant trees) was calculated as the Moran Index, \textit{I} applied to the start of fruiting dates. Since the date of start of fruiting corresponds to a circular variable, I followed [@schmal2017moran] to adapt existing functions of the *ape* package [@ape] to circular variables. Moran's \textit{I} was therefore defined as

@@

$I = \frac{1}{\sum_{ij}{w_{ij}}}\frac{\sum_{ij}w_{ij}d(x_i,\bar{t})d(x_j,\bar{t})}{N^{-1}\sum_i d(x_i,\bar{t})^2}$

@@

<!-- $C = \frac{(N-1)}{2} \frac{\sum_{i}\sum_{j} w_{ij}d(t_i,t_j)^2}{\sum_{i,j}w_{ij}\sum_i d(x_i,\bar{t})^2}$ [@geary1954contiguity] -->

where $N$ is the number of trees, $t$ is the fruiting date (in radian), $d(t_1,t_2) = atan2(sin(t_1 - t_2), cos(t_1 - t_2))$, $\bar{t} = atan2(1/N\sum_isin(t_i), 1/N\sum_icos(t_i))$. I used the inverse of the Euclidean distance between tree locations as weights. Values of $C$ close to 0 indicate absence of spatial autocorrelation. Values close to -1 indicate negative spatial autocorrelation. Values close to 1 indicate positive spatial autocorrelation.

@@

For each simulation, I verified whether a routine behaviour emerged and paralleled the variations in tree distribution. I characterised routine holding on the notion of conditional entropy following [@riotte2016randomness] using the series of the id of trees targeted during five supplementary seasons in which dispersal could not occur. I computed the routine possibly up to only a first order conditional entropy, as computing routine indices is computationally demanding and a first order is expected to be the most informative to describe a perfect traplining behaviour [@riotte2016randomness]. The routine index varies from 0 (no routine) to 1 (perfect traplining).

@@

To investigate whether different levels of cognition eventually led to different distribution patterns, I considered five forager types differing by the proportion of trees for which the location and phenology were known (from 0 to 100%, at a step of `r 100/length(spatialKnowledge)`). For each forager, `r numberRepetitions` simulations were performed for a given condition (i.e. knowledge rate), in which the forager foraged individually `r cycleLimitNumber` seasonal periods.

@@

To investigate whether the engineered distribution of trees could in turn affect the foraging benefits purported by cognition, I used the final tree distributions of simulations computed to answer [question 1][Q1. Does resource spatiotemporal distribution vary with cognition level?]). For each of these tree distributions, in three different simulation runs, I let forage a forager endowed with no, intermediate (i.e. half of trees known), or full (all trees known, omniscience/prescience, hereafter only omniscience) spatiotemporal knowledge. In these simulations, no seed-dispersal could occur. I measured the foraging efficiency after the forager foraged for `r cycleLimitNumber/5` full seasons and compared them across environments and forager abilities. Foraging efficiency was computed as the cumulated quantity of eaten food, over the cumulated distance travelled. I verified that at the end of the simulation that foraging efficiency stabilised, which meant that I reached a foraging equilibrium. I considered that foraging efficiency stabilised when the foraging efficiency sampled after each moving bout during the fourth and fifth quintile of simulation times did not differ of more than `r percentToAssumeDifference`% (reference is the minimum efficiency in the fourth or the fifth quintile of simulation times). I repeated the simulations for a given condition (i.e. knowledge rate) `r numberRepetitions` times.

@@

During the model construction, I noticed that two mechanisms could amplify the phenomenon of route formation, in addition to the use of memory: space competition between trees and sequential targeting of all trees seen *en route* to the target. To better quantify the effect of those two forces, I considered two other sets of `r numberRepetitions` simulations each.

|   To assess the effect of competition for space between trees, I compared the final tree spatiotemporal patterns when an omniscient forager (all tree locations and phenology known) was let foraging in the same setting as in question 1, with the only difference that the exclusive lifespace between trees could be of 5 (small lifespace), 45 (intermediate lifespace) or 85% (large lifespace) of the map surface, so that the radius of this exclusive circular lifespace was always inferior to the perception range (`r round(0.85 * sqrt(mapSize * mapSize / (numberTrees * pi)), digits = 2)` su vs `r round(perceptualRange, digit=2)` su).

@@

```{r figEngineering, fig.width=19.25, fig.height=19.25, include=TRUE, warning = FALSE, fig.cap="Seed-dispersing foragers can shape resource distribution | The plots show the results of 200 simulations when a seed-dispersing forager with varying level of cognition (i.e. proportion of tree locations and associated phenology known), moving choice rules or space competition intensities between trees (the two latter being simulated with an omniscient forager; left to right) was let foraging for a long time. The resulting distribution pattern was assessed by its patchiness (the tendency of trees to be aggregated into patches; larger values indicate highly aggregated trees), its alignement (the tendency of trees to form linear geometries; larger values indicate the presence of linear geometries) and spatial autocorelation in fruiting date (the tendency of trees to have fruiting dates closer/farther in time to their spatial neighbours; values close to 0 indicate negative spatial autocorrelation, values close to 1 indicate no spatial autocorrelation, values above 1 indicate positive spatial autocorrelation). The emerging movement behaviour of the forager was described in terms of routine behaviour (the tendency of the forager to target trees in a predictible way; values close to 0 indicate no predictible movement, values close to 1 indicate highly predictible movements)."}

@@

```{r figEfficiency, fig.width=8, fig.height=6, include=TRUE, warning = FALSE, fig.cap=" Relative foraging benefits of cognition are sensitive to the emerging resource distribution | The 'initial condition' of resource distribution was taken from the simulations in which a null, intermediate or omniscient forager was first let foraging. Then, those three forager types were let foraging (separately) without dispersal and foraging efficiency was measured (cumulated food eaten over cumulated travelled distance)."}

@@

## Variations in routine behaviour when no dispersal occurs
